Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-10-26 ms.tex
Adding short paragraph in WKB appendix that explains why the asymmetric piece of the solution is missing

Let us define the damping rate to be $\gamma \equiv i\omega$, which is real and positive for damped solutions. At the eigenvalues $\gamma=\gamma_{nm}$, the response $J(n, \sigma, \omega)$ is resonant. Near these $\gamma_{nm}$ poles, the eigenfunctions will have the form 
\be \label{eq:jnsigmaomega_approx}
J(n,\sigma,-i\gamma) & \simeq \frac{ J_{nm}(\sigma) }{\gamma_{nm} - \gamma} + {\rm terms\ which\ vary\ slowly\ in\ \gamma}.
\ee
From Eq. \ref{eq:jrsigmat_expansion}, the contribution of a single pole to the total solution is
\be
J_{nm}(r,\sigma,t) & = &  -j_0(\kappa_n r)  \int \frac{d\gamma}{2\pi i} e^{-\gamma t}J(n,\sigma,-i\gamma)
\\ & \simeq & j_0(\kappa_n r)  \int \frac{d\gamma}{2\pi i} e^{-\gamma t} \frac{ J_{nm}(\sigma) }{\gamma - \gamma_{nm}}.
\ee
We close the integral along the imaginary $\gamma$-axis  with a semi-circle at infinity and apply the residue theorem. The contribution from the semi-circle is zero, so we obtain
\be
J_{nm}(r,\sigma,t) & \simeq & j_0(\kappa_n r) J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Summing over all spatial modes $n$ and over all eigenmodes $m$ for a given $n$, we obtain
\be \label{eq:Jrsigmat}
J(r,\sigma,t) & = & \sum_{n=1}^\infty j_0(\kappa_n r)  \sum_{m=1}^{\infty} J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
This ansatz captures the contributions from $n \times m$ simple poles, which seems to agree with Monte Carlo results for $x_s=0$ based on numerical calculations of this result. However, the analysis presented in Appendix \ref{app:wkb} suggests the ansatz is incomplete in the case where $x_s \neq 0$ and the spectrum is asymmetric.

@@

A primary goal of this work is to present a solution for mean intensity ($J$) and flux ($H = F/4\pi$) of photons near line-center frequency $\nu_0$ in a uniform sphere. We have generalized a spherically symmetric solution derived by \cite{2006ApJ...649...14D} (called $H_0$ here) to allow a monochromatic source of photons with frequencies away from line center. We have made a correction to their incorrect $J=0$ surface boundary condition that allows the proper condition $J=\sqrt{3}H$ to be satisfied at the surface of the sphere. This introduces a term $J_{\rm bc}$ which is solved using a continuous Fourier expansion in frequency. Enforcement of the correct boundary term leads naturally to expansion of the spatial dependence in terms of spherical Bessel functions. The continuous expansions are discretized and the Fourier coefficients solved for. The resulting flux correction, $H_{\rm bc}$, scales as $(a\tau_0)^{-1/3}$. Thus, for large $\tau_0$, only a small correction to $H_0$ is needed, while larger errors are present in calculations performed at lower $\tau_0$. Since the Laplacian form of the differential equation is only attainable by approximating the line profile as $\phi \approx a/(\pi x^2)$, our solutions do not model areas where the Doppler core of the \lya line has a significant effect on the scattering of photons. Because the peak of the spectral energy distribution of escaping photons is $x_0 = (a\tau_0)^{1/3}$, calculations performed at low $\tau_0$ are inaccurate due to the close proximity of the spectral peak and the Doppler core of the line. This effect is especially prevalent for larger H gas temperature, as the boundary between the Doppler core and line wing extends further out as $T$ increases.

@@

We have examined previous solutions to \lya transfer in the limit of large optical depth, noting that the separation of variables and treatment of the boundary condition in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others does not produce a valid solution to the transfer equation. Here, we have derived the solution in spherical geometry with an appropriate treatment of the surface boundary condition as a frequency-dependent quantity, examining the spectral and time dependence of the resulting equations. The key result is that the errors in the previously-cited works have been quantified via a corrective boundary term, $H_{\rm bc}$, which resolves an excess in flux at the spectral peak and a deficit in the line wing of the calculated spectrum of \lya radiation as compared with Monte Carlo. The size of $H_{\rm bc}$ diminishes at larger $\tau_0$, following a $(a\tau_0)^{-1/3}$ scaling. Introducing time-dependence to the transfer equation leads to another solution which is solved numerically with an eigenfunction expansion. We demonstrate that it agrees with the steady-state spectrum for $x_s=0$ when integrated over time, though its rate of numerical convergence is slow and requires a sum over many modes to become accurate. This treatment is incomplete for emission away from line center ($x_s\neq0$), however, since any terms that would allow asymmetry in the spectrum are not present in the sum. The time-dependent solution is utilized to create wait-time distributions for photons escaping the sphere of optically-thick hydrogen gas, which can be sampled in lieu of direct Monte Carlo simulations that would ordinarily be computationally demanding. We compare the calculations from the time-dependent solution with Monte Carlo for a sample of $\tau_0$, noting general agreement in the resulting escape time distributions. In future work, this will be used to accelerate Monte Carlo \lya transfer at large optical depths with potential applications in radiation hydrodynamic simulations of the atmospheres of exoplanets.

@@

We start with the differential equation Eq. \ref{eq:wkb_differential_eqn}. Near line center, this has two independent solutions which can be written
\be
J(n,\sigma,s) & = & a_1 \left( \frac{\sigma}{\sigma_{\rm tp}} \right)^{1/2} J_{-3/4}\left( k_c \left[ \frac{\sigma}{\sigma_{\rm tp} }\right]^{2/3} \right)
\nonumber \\ &  + &  a_2 \left( \frac{\sigma}{\sigma_{\rm tp}} \right)^{1/2} J_{3/4}\left( k_c \left[ \frac{\sigma}{\sigma_{\rm tp} }\right]^{2/3} \right),
\ee
where $J_\alpha(...)$ are Bessel functions of the first kind, $a_1$ and $a_2$ are normalization constants, $\sigma_{tp}$ is given in Eq. \ref{eq:sigma_tp}, and $k_c=3\kappa_n \Delta \sigma_{\rm tp}/2k$. The solution far from line center is 
\be
J(n,\sigma,s) & = & a_3 \left( \frac{\sigma}{\sigma_{\rm tp}} \right)^{1/6} Ai\left(k_c^{2/3} \left[ \left( \frac{\sigma}{\sigma_{\rm tp}} \right)^{2/3} - 1 \right]\right),
\ee
where $Ai(...)$ is the Airy function of the first kind. Again, $a_3$ is a normalization constant. When $\sigma_s=0$, we can enforce the discontinuity in Eq. \ref{eq:matching_condition_2} to find 
\be
a_2 & = & - \frac{\Gamma(7/4) \sigma_{\rm tp}}{ 2^{1/4} k_c^{3/4}} \frac{6^{1/2}}{8} n^2 \frac{E}{kR^3}.
\ee
The discontinuity does not set the value of $a_1$ as that solution does not have a term linear in $\sigma$ when the source is at $\sigma_s=0$. However, where both functions have short-wavelength behavior we may show that the solutions near and far from $\sigma=0$ have the same approximate form in terms of sines and cosines. This matched asymptotic expansion allows us to relate the normalization coefficients $a_1$ and $a_3$. The matching conditions are
\be
\frac{a_1}{a_2} & = & \frac{ \sin \left( \frac{\kappa_n \sigma_{\rm tp} \Delta}{k} + \frac{\pi}{8} \right) }{ \sin \left(  \frac{\kappa_n \sigma_{\rm tp} \Delta}{k} - \frac{\pi}{8} \right) }
\\
\frac{a_3}{a_2} & = & \frac{ k_c^{-1/3} }{ \sin \left(  \frac{\kappa_n \sigma_{\rm tp} \Delta}{k} - \frac{\pi}{8} \right) }.
\ee
By setting the to denominator of this expression equal to zero and solving it for the eigenfrequency $\gamma_{nm}$ contained in $\sigma_{tp}$, we find the dispersion relation in Eq. \ref{eq:gamma_nm}.

>>>>

Let us define the damping rate to be $\gamma \equiv i\omega$, which is real and positive for damped solutions. At the eigenvalues $\gamma=\gamma_{nm}$, the response $J(n, \sigma, \omega)$ is resonant. Near these $\gamma_{nm}$ poles, the eigenfunctions will have the form 
\be \label{eq:jnsigmaomega_approx}
J(n,\sigma,-i\gamma) & \simeq \frac{ J_{nm}(\sigma) }{\gamma_{nm} - \gamma} + {\rm terms\ which\ vary\ slowly\ in\ \gamma}.
\ee
for which an application of the residue theorem gives
\be
J_{nm}(r,\sigma,t) & \simeq & j_0(\kappa_n r) J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Summing over all spatial modes $n$ and over all eigenmodes $m$ for a given $n$, we obtain
\be \label{eq:Jrsigmat}
J(r,\sigma,t) & = & \sum_{n=1}^\infty j_0(\kappa_n r)  \sum_{m=1}^{\infty} J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
This ansatz captures the contributions from $n \times m$ simple poles, which seems to agree with Monte Carlo results for $x_s=0$ based on numerical calculations of this result. However, the analysis presented in Appendix \ref{app:wkb} suggests the ansatz is incomplete in the case where $x_s \neq 0$ and the spectrum is asymmetric.

@@

A primary goal of this work is to present a solution for mean intensity ($J$) and flux ($H = F/4\pi$) of photons near line-center frequency $\nu_0$ in a uniform sphere. We have generalized a spherically symmetric solution derived by \cite{2006ApJ...649...14D} (called $H_0$ here) to allow a monochromatic source of photons with frequencies away from line center. We have made a correction to their incorrect $J=0$ surface boundary condition that allows the proper condition $J=\sqrt{3}H$ to be satisfied at the surface of the sphere. This introduces a term $J_{\rm bc}$ which is solved using a continuous Fourier expansion in frequency. Enforcement of the correct boundary term leads naturally to expansion of the spatial dependence in terms of spherical Bessel functions. The continuous expansions are discretized and the Fourier coefficients solved for. The resulting flux correction, $H_{\rm bc}$, scales as $(a\tau_0)^{-1/3}$. Thus, for large $\tau_0$, only a small correction to $H_0$ is needed, while larger errors are present in calculations performed at lower $\tau_0$. Since the Laplacian form of the differential equation is only attainable by approximating the line profile as $\phi \approx a/(\pi x^2)$, our solutions do not model areas where the Doppler core of the \lya line has a significant effect on the scattering of photons. Because the peak of the spectral energy distribution of escaping photons is $x_{\rm peak} = (a\tau_0)^{1/3}$, calculations performed at low $\tau_0$ are inaccurate due to the close proximity of the spectral peak and the Doppler core of the line. This effect is especially prevalent for larger H gas temperature, as the boundary between the Doppler core and line wing extends further out as $T$ increases.

@@

We have examined previous solutions to \lya transfer in the limit of large optical depth, noting that the separation of variables and treatment of the boundary condition in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others does not produce a valid solution to the transfer equation. Here, we have derived the solution in spherical geometry with an appropriate treatment of the surface boundary condition as a frequency-dependent quantity, examining the spectral and time dependence of the resulting equations. The key result is that the errors in the previously-cited works have been quantified via a correction term, $H_{\rm bc}$, which resolves an excess in flux at the spectral peak and a deficit in the line wing of the calculated spectrum of \lya radiation as compared with Monte Carlo. The size of $H_{\rm bc}$ diminishes at larger $\tau_0$, following a $(a\tau_0)^{-1/3}$ scaling. Introducing time-dependence to the transfer equation leads to another solution which is solved numerically with an eigenfunction expansion. We demonstrate that it agrees with the steady-state spectrum for $x_s=0$ when integrated over time, though its rate of numerical convergence is slow and requires a sum over many modes to become accurate. This treatment is incomplete for emission away from line center ($x_s\neq0$), however, since any terms that would allow asymmetry in the spectrum are not present in the sum. The time-dependent solution is utilized to create wait-time distributions for photons escaping the sphere of optically-thick hydrogen gas, which can be sampled in lieu of traditional Monte Carlo simulations that would be computationally demanding. We compare the calculations from the time-dependent solution with Monte Carlo for a sample of $\tau_0$, noting general agreement in the resulting escape time distributions. In future work, this will be used to accelerate Monte Carlo \lya transfer at large optical depths with potential applications in radiation hydrodynamic simulations of the atmospheres of exoplanets.

@@

We start with the differential equation Eq. \ref{eq:wkb_differential_eqn}. Near line center, this has two independent solutions which can be written
\be \label{eq:wkb_J}
J(n,\sigma,s) & = & a_1 \left( \frac{\sigma}{\sigma_{\rm tp}} \right)^{1/2} J_{-3/4}\left( k_c \left[ \frac{\sigma}{\sigma_{\rm tp} }\right]^{2/3} \right)
\nonumber \\ &  + &  a_2 \left( \frac{\sigma}{\sigma_{\rm tp}} \right)^{1/2} J_{3/4}\left( k_c \left[ \frac{\sigma}{\sigma_{\rm tp} }\right]^{2/3} \right),
\ee
where $J_\alpha(...)$ are Bessel functions of the first kind, $a_1$ and $a_2$ are normalization constants, $\sigma_{tp}$ is given in Eq. \ref{eq:sigma_tp}, and $k_c=3\kappa_n \Delta \sigma_{\rm tp}/2k$. The solution far from line center is 
\be
J(n,\sigma,s) & = & a_3 \left( \frac{\sigma}{\sigma_{\rm tp}} \right)^{1/6} Ai\left(k_c^{2/3} \left[ \left( \frac{\sigma}{\sigma_{\rm tp}} \right)^{2/3} - 1 \right]\right),
\ee
where $Ai(...)$ is the Airy function of the first kind. Again, $a_3$ is a normalization constant. When $\sigma_s=0$, we can enforce the discontinuity in Eq. \ref{eq:matching_condition_2} to find 
\be
a_2 & = & - \frac{\Gamma(7/4) \sigma_{\rm tp}}{ 2^{1/4} k_c^{3/4}} \frac{6^{1/2}}{8} n^2 \frac{E}{kR^3}.
\ee
The discontinuity sets the value of $a_2$, and the corresponding term in Eq. \ref{eq:wkb_J} is non-resonant at $\sigma_s=0$. When $\sigma_s \neq 0$, however, this component contains the piece of the solution that is asymmetric about the line center. Notably, this piece of the solution is not included in Eq. \ref{eq:Jrsigmat} and thus our treatment using the residue theorem does not capture contributions from emission away from line center.

%The discontinuity does not set the value of $a_1$ as that solution does not have a term linear in $\sigma$ when the source is at $\sigma_s=0$. 
Where both functions have short-wavelength behavior, we may show that the solutions near and far from $\sigma=0$ have the same approximate form in terms of sines and cosines. This matched asymptotic expansion allows us to relate the normalization coefficients $a_1$ and $a_3$. The matching conditions are
\be
\frac{a_1}{a_2} & = & \frac{ \sin \left( \frac{\kappa_n \sigma_{\rm tp} \Delta}{k} + \frac{\pi}{8} \right) }{ \sin \left(  \frac{\kappa_n \sigma_{\rm tp} \Delta}{k} - \frac{\pi}{8} \right) }
\\
\frac{a_3}{a_2} & = & \frac{ k_c^{-1/3} }{ \sin \left(  \frac{\kappa_n \sigma_{\rm tp} \Delta}{k} - \frac{\pi}{8} \right) }.
\ee
By setting the to denominator of this expression equal to zero and solving it for the eigenfrequency $\gamma_{nm}$ contained in $\sigma_{tp}$, we find the dispersion relation in Eq. \ref{eq:gamma_nm}.
